---
title: "Mapping free text to ontology terms with map_concepts"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Mapping free text to ontology terms with map_concepts}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

```{r setup, echo=FALSE, message=FALSE}
library(biocEDAM)
```

# Introduction

`map_concepts()` and `map_concepts_edam()` translate free text — such as a
dataset description, abstract, or workflow summary — into structured ontology
term mappings using a two-stage LLM pipeline backed by the
[EBI OLS4](https://www.ebi.ac.uk/ols4) ontology search service.

**Stage 1** extracts a list of biological, medical, or computational concepts
from the input text using a plain LLM call (no external tools).

**Stage 2** looks up each concept individually in OLS4 via a fresh LLM
conversation, returning a term label.  R code then resolves that label to a
canonical IRI via the OLS4 REST API, keeping the LLM out of IRI assignment
entirely.  Each concept receives its own fresh conversation so that context
never accumulates across concepts — a key safeguard against the hallucinated
IRIs that arise when a single long conversation handles many tool calls.

# Requirements

Two external dependencies are needed for live use.  When either is absent the
code in this vignette falls back automatically to stored mock results so the
vignette still builds and displays representative output.

**1. Anthropic API key** — set before calling any `map_concepts` function:

```{r auth, eval=FALSE}
Sys.setenv(ANTHROPIC_API_KEY = "your-key-here")
```

**2. OLS4 MCP bridge** — `npx` and `mcp-remote` must be on the system PATH.
Install once with:

```{bash mcp_install, eval=FALSE}
npm install -g mcp-remote
```

# map_concepts: multi-ontology mapping

## A bioinformatics workflow description

`map_concepts()` searches across all ontologies available in OLS4, selecting
the most appropriate one for each concept.  The `max_concepts` parameter caps
the number of OLS4 lookups, which is useful for long texts.

```{r wf_text}
wf_text <- paste(
    "The workflow for Illumina-sequenced ARTIC data builds on the RNASeq",
    "workflow for paired-end data using the same steps for mapping and variant",
    "calling, but adds extra logic for trimming ARTIC primer sequences off",
    "reads with the ivar package."
)
```

When an API key and MCP bridge are available, run live:

```{r mc_workflow}
if (interactive())
    wf <- map_concepts(wf_text, max_concepts = 8)
```

The vignette uses a pre-computed result:

```{r mc_workflow_mock, message=FALSE}
wf <- readRDS(system.file("extdata", "mock_wf_map_concepts.rds",
                           package = "biocEDAM"))
as.data.frame(wf)[, c("input_text", "term_label", "obo_id", "ontology")]
```

Note that `deduplicate = TRUE` (the default) collapses rows that resolve to the
same IRI, recording all source concepts in the `input_text` field separated by
`"; "`.

## A cardiovascular genomics abstract

For a longer, more heterogeneous text `map_concepts()` spans multiple
ontologies — MONDO for diseases, EFO for study designs, EDAM for
computational methods:

```{r cvd_text}
cvd_text <- paste(
    "The Cardiovascular Disease working group considered early-onset coronary",
    "artery disease, stroke, atrial fibrillation, congestive heart failure and",
    "type 2 diabetes. Genome-wide association studies provide a powerful tool",
    "to identify common variants. Comprehensive molecular phenotyping is",
    "performed using genomic, transcriptomic, proteomics, and metabolomic",
    "approaches."
)
```

```{r mc_cvd}
if (interactive())
    cvd <- map_concepts(cvd_text, max_concepts = 12)
```

```{r mc_cvd_mock, message=FALSE}
cvd <- readRDS(system.file("extdata", "mock_cvd_map_concepts.rds",
                            package = "biocEDAM"))
as.data.frame(cvd)[, c("input_text", "term_label", "obo_id", "ontology")]
```

# map_concepts_edam: EDAM-only mapping

`map_concepts_edam()` restricts both the LLM search and the OLS4 REST label
resolution to the EDAM ontology.  EDAM has four sub-trees:

| Sub-tree | Scope |
|---|---|
| `topic` | Research areas and scientific domains |
| `operation` | Computational steps and analytical methods |
| `data` | Data types and information entities |
| `format` | File formats and data exchange standards |

```{r mce}
if (interactive())
    wf_edam <- map_concepts_edam(wf_text, max_concepts = 8)
```

```{r mce_mock, message=FALSE}
wf_edam <- readRDS(system.file("extdata", "mock_wf_map_concepts_edam.rds",
                                package = "biocEDAM"))
as.data.frame(wf_edam)[, c("input_text", "term_label", "obo_id")]
```

## Scope limitations

EDAM does not cover clinical phenotypes, diseases, spatial statistics, or
general study designs.  For the cardiovascular text above, `map_concepts_edam()`
would return very few rows — only the omics method concepts (genomics,
transcriptomics, proteomics) have good EDAM coverage.  Use `map_concepts()` for
mixed or clinically-oriented texts.

# Parameters

| Parameter | Default | Purpose |
|---|---|---|
| `max_concepts` | `Inf` | Cap Stage 2 lookups; use 8–15 for long texts |
| `deduplicate` | `TRUE` | Collapse rows with identical IRIs |
| `definition` | `FALSE` | Fetch OLS4 definitions (one extra REST call per term) |
| `label_match` | `FALSE` | Add `llm_label`/`label_match` diagnostic columns |
| `ontology_filter` | `NULL` | Force REST search to a specific ontology |
| `tools` | `ols4_mcp_tools()` | Pre-loaded MCP tools; reuse to avoid bridge restarts |

Pre-loading the MCP tools is useful when calling `map_concepts` repeatedly:

```{r preload}
if (interactive()) {
    tls <- ols4_mcp_tools()
    r1  <- map_concepts(text1, tools = tls)
    r2  <- map_concepts(text2, tools = tls)
}
```

# Curation

All outputs require human review before use.  Common failure modes include:

- A plausible term label that resolves to an unrelated IRI via OLS4 search
- A concept mapped to an ontology whose scope does not match (e.g. a disease
  concept mapped to a GO biological process)
- A spurious EDAM term returned for a concept outside EDAM's scope

Setting `definition = TRUE` adds authoritative OLS4 definitions, which makes
it easier to judge whether a mapping is appropriate:

```{r def}
if (interactive()) {
    wf_def <- map_concepts_edam(wf_text, max_concepts = 8, definition = TRUE)
    print(wf_def[, c("term_label", "definition")])
}
```

Inspect `term_label` against `input_text` for each row and discard implausible
mappings before treating results as authoritative.

# Inspecting prompts

All prompts used internally can be inspected and customised via `read_prompt()`:

```{r prompts}
cat(read_prompt("extract_concepts.txt"))
```

```{r prompts2}
cat(read_prompt("lookup_edam_concept.txt"))
```

# Session information

```{r si}
sessionInfo()
```
